Fitting metabolic models to dissolved oxygen data: The Estuarine BAyesian Single-station Estimation method

### 2023-08-01 manu-draft.qmd
added info about missing data

load(file = here('tabs/ebaseiotab.RData'))
load(file = here('tabs/apacmptab.RData'))
load(file = here('data/apasumdat.RData'))
```

@@

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \delta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $R$, and $b$. All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.

>>>>

load(file = here('tabs/ebaseiotab.RData'))
load(file = here('tabs/apacmptab.RData'))
load(file = here('data/apasumdat.RData'))
load(file = here('data/misdat.RData'))
```

@@

where $N(\alpha, \beta)$ indicates a normal distribution with mean $\alpha$ and standard deviation $\beta$ and $T(\gamma, \theta)$ is 1 between $\gamma$ and $\theta$ and 0 elsewhere; $T$ has the effect of truncating the normal distribution to minimum and maximum values. All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase. Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.502 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ (\~twice the default mean). Initial development of EBASE showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates. Examples of the prior distributions in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] are shown in @fig-priorplot.

@@

```{r}
wqmis <- misdat$Temp # same percentages across all wq parms
mtmis <- misdat$totpar # same percentages across all met parms
```

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $R$, and $b$ (@eq-ebase) was conducted using a synthetic time series with actual data as inputs. These assessments evaluated if 1) EBASE estimates similar values for the known parameters in the synthetic time series, and 2) how the estimated values change with different model configurations. The synthetic time series used inputs from one year of continuous data from Apalachicola Bay, Florida. The input data were observations from the year 2021 at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, --84.8802$^{\circ}$ W) and included 15-minute observations of water temperature and salinity. Continuous meteorological observations from the nearby East Bay station (29.7694$^{\circ}$ N, --84.8815$^{\circ}$ W, \~10 km north of Cat Point) were combined with the water quality data at the same time step and included air temperature, wind speed, and PAR. Missing observations in the water quality and meteorological data (`r wqmis`% and `r mtmis`% of all observations in each dataset, respectively) were filled using autoregressive modelling of the actual parameters versus time to create a complete dataset from January 1st to December 31st. A synthetic oxygen time series was then generated using the inputs from Apalachicola Bay ([@fig-synapa]a) and @eq-ebase as a forward calculating model with user-defined values for $a$, $R$, and $b$. All metabolic rates for production, respiration, and gas exchange ([@fig-synapa]b and c) were also estimated with the synthetic time series. The only parameter that could not be created as a continuous time series was $b$, which was set to 0.251 $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$ [@Wanninkhof14]. Accordingly, the synthetic dissolved oxygen time series and the observed water quality and meteorological data were used with EBASE to determine if reasonable estimates for the known parameters and metabolic rates were recovered.
